---
title: "Multinomial"
format:
  html:
    toc: true
    html-math-method: katex
    css: style.css  
    theme:
      light: cosmo
      dark: [cosmo, theme-dark.scss]
execute: 
   echo: false
editor_options: 
  chunk_output_type: console
---

```{=html}
<style>
.table-hover > tbody > tr:hover { 
  background-color: #f4f442;
}
</style>
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(here)
source(here("source","get_lib.R"))
```

-   Reading
    -   Ch.3
        -   skip 3.4.3, 3.5 (page )


# Multinomial probability distribution

-   Suppose `Y` is categorical response random variable and we have `n` trials, $Y_1..Y_n$
-   $N_j$ counts the number of trials responding with category `j`.
    -   $N_j = \sum_{i=1}^n I(Y_i=j)$
-   parameters:
    -   number of category, `j=1,...J`
    -   probability associated with each category $\pi_j = P(Y=j)$
-   It is distribution, so

$$\sum_{j=1}^J \pi_j = 1$$

-   Then, PMF of observing a particular set of counts $n_1,...n_j$ is

$$P(N_1 = n_1,....N_j = J) = \frac{n!}{\prod_{j=1}^{J}n_j!}\prod_{j=1}^{J}\pi_j^{n_j}$$

This is known as the `multinomial probability distribution`

-   We use MLE to obtain estimates of $\pi_1,...\pi_J$ which is $\hat{\pi} = \frac{n_j}{n}$, which is the observed proportion for each category.


# One Multinomial distribution

-   This is different from `multinomial distribution`.
-   Well... here is an example [LINK](https://docs.google.com/spreadsheets/d/1YyO_sX1hIn7siATwF7L-mE8OfHktGX2w/edit?gid=139180726#gid=139180726)

```{r}
#| echo: true
pi.ij <- c(0.2, 0.3, 0.2, 0.1, 0.1, 0.1)
pi.table <- array(data = pi.ij , dim = c(2,3), dimnames = list(X
= 1:2, Y = 1:3))

pi.table %>% kable("html") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r}
#| echo: true
set.seed(9812)  

save <- rmultinom(n = 1, size = 1000, prob = pi.ij)  

c.table1 <- array(data = save , dim = c(2,3), dimnames = list(X
= 1:2, Y = 1:3))

c.table1/sum(c.table1)

rowSums(c.table1)
```

# Product Multinomial distribution

-   For the I multinomial (i.e., product multinomial) setting, we again simulate a sample for a 2 × 3 contingency table (thus, I = 2).
-   With this model, we need to draw samples of fixed size separately for each row.
  - Question is can we get joint probability using this method?
-   Here is an example from book

## This part is preparing the data set part

```{r}
#| echo: true
#calculate conditional probability 
pi.cond <- pi.table/rowSums(pi.table)
pi.cond %>% kable("html") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))

#simulate based on the conditional probability
#notice the sample size is different from c.table1
#suppose that we sampled 400 and 600.  This is just what was 
#observed or we selected this
set.seed(8111) 
save1 <- rmultinom(n = 1, size = 400, prob = pi.cond[1,]) 
save2 <- rmultinom(n = 1, size = 600, prob = pi.cond[2,])

c.table2 <- array(data = c(save1[1], save2[1], save1[2],
save2[2], save1[3], save2 [3]), dim = c(2,3), dimnames = list(X = 1:2, Y = 1:3))

c.table2 %>% kable("html") %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))
```

-   `Can we estimate the joint from this table?` is the question.

    -   Answer is, `yes`

```{r}
    c.table2/rowSums(c.table2)
```

-   Now, we have joint and marginal, can we test independence?

# Testing for independence

- See page 148

$$H_0: P(X=x_i,Y=y_j) = P(x_i)\cdot P(y_j)$$ $$H_a: P(X=x_i,Y=y_j) \not= P(x_i)\cdot P(y_j)$$ Using the book notation in p147,

$$H_0: \pi_{ij}=\pi_{i+}\pi_{+j} \text{  for each i,j}$$ $$H_0: \pi_{ij}\not=\pi_{i+}\pi_{+j} \text{  for some i,j}$$

### Compare test statistic and $\chi^2$

-   This is called `Pearson Chi-square test`
-   Under product multinomial model (see page 147), this is how you compute the test statistics

$$\text{X}^2 = \frac{\text{difference between joint and product of marginal}}{\text{product of marginal}}$$

-   Above statistic has an approximate $\chi^2$ distribution with certain degrees of freedom

-   After you set your type I error, $\alpha$, you get your crtical value and compare that critical value with your observed test statistic $\text{X}^2$

### LR test

-   This is called LRT test
-   The likelihood ratio test can also be conducted.
    -   The joint probability can be computed in two different ways
        -   From contingency table, just calculate it.
        -   And using the marginal probabilities, multiply them.
    -   Then, using equation 3.2 in page 143, you can calculate the likelihood of both and form `the likelihood ratio`

$$\lambda = \frac{\text{the MLF estimated using joint probabilites}}{\text{the MLF estimated using joint probabilities estimated as the product of marginal}}$$

And $-2\text{log}(\lambda)$, also follows $\chi^2$ distribution with certain degrees of freedom. (this sounds like a magic.. to me..)

### Note

-   Both LRT and the Pearson chi-square test generally give similiar result in large samples. (see page 147)
    -   common criteria $\frac{\text{product of marginal counts}}{\text{total count}} > 1$ or $>5$ for all cells of the contingency table

# Example of Testing Independnece 

-   We would like to determine if `bloating severity` is related to the `type of fiber`.

```{r}
diet <- read.csv(here("data","Fiber.csv"), stringsAsFactors = TRUE)

# Match order given in table
diet$fiber <- factor(x = diet$fiber, levels = c("none", "bran", "gum", "both"))
diet$bloat <- factor(x = diet$bloat, levels = c("none", "low", "medium", "high"))

diet.table <- xtabs(formula = count ~ fiber + bloat, data = diet)
diet.table
```

```{r}
# Hypothesis tests for independence
ind.test <- chisq.test(x = diet.table, correct = FALSE)
ind.test
```

-   Note that we include the `correct = FALSE` argument value in `chisq.test()` to prevent a continuity correction from being applied.
    -   Please see Section 1.2.3 for why we avoid these corrections.

```{r}
ind.test$expected
```

