---
title: "Ordinal response regression models"
format:
  html:
    toc: true
    html-math-method: katex
    css: style.css  
    theme:
      light: cosmo
      dark: [cosmo, theme-dark.scss]
execute: 
   echo: false
editor_options: 
  chunk_output_type: console
---

```{=html}
<style>
.table-hover > tbody > tr:hover { 
  background-color: #f4f442;
}
</style>
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(here)
source(here("source","get_lib.R"))
```

-   Reading
    -   Ch.3.4
        -   skip 3.4.3, 3.5 

# Proportions odds model

- ordinal (i.e., ordered like Likert scale) or 

## Functional form 

- (p170)
- The model is the log-odds of cumulative probabilities, called `cumulative logits`

$$\text{logit}(P(Y \le j)) = \text{log}\frac{P(Y \le j)}{1-P(Y \le j)}$$
In particular, the `proportional odds model` is assumes the probabilities changes linear as the explanatory variable chagne and the also the slope of this relationship is the same regardless of the category `j`. (see page 170)


$$\text{logit}(P(Y \le j)) = \beta_{j0} + \beta_{1}x_1 .. + \beta_{p}x_p$$
- Notice the subscript `j` exists only for the constant term. This is the simplified version of the nominal response model.

- The model assumes that the effects of the explanatory variables are the same regardless of which cumulative probabilities are used to from the `log odds`

- For a fixed `j`, increasing $x_j$ by `c` units changes every log-odds by $c\beta_r$ when holding other explanatory variables constant.

## Probability 

$$\pi_1 = P(Y \le 1) - P(Y \le 0) = \frac{\text{exp}(\beta_{j0} + \beta_{1}x_1 .. + \beta_{p}x_p)}{1+\text{exp}(\beta_{j0} + \beta_{1}x_1 .. + \beta_{p}x_p)}$$
$$\pi_J = P(Y \le J) - P(Y \le J-1) = \frac{\text{exp}(\beta_{J-1,0} + \beta_{1}x_1 .. + \beta_{p}x_p)}{1+\text{exp}(\beta_{J-1,0} + \beta_{1}x_1 .. + \beta_{p}x_p)}$$


# Wheat problem

- Now assuming `scab (Y=1) < sprout (Y=2) < healthy (Y=3)` and fit proportional odds model to the same data

```{r}
wheat <- read.csv(here("data","Wheat.csv"))
glimpse(wheat)
wheat$type <- as.factor(wheat$type)
levels(wheat$type)

#reorder the response
wheat$type.order <- factor(wheat$type , levels = c("Scab",
"Sprout", "Healthy"))

levels(wheat$type.order)

```

## fit the model 

- `method = "logistic" in polr()` argument instruct R to use the logit transformation on the cumulative probabilities.

```{r}
mod.fit.ord <- polr(formula = type.order ~ class + density +
hardness + size + weight + moisture , data = wheat , method = "logistic")

# latex_equation <- extract_eq(mod.fit.ord)
# print(latex_equation)
```

$$
\begin{aligned}
\log\left[ \frac { P( \operatorname{type.order}  \leq  \operatorname{Scab} ) }{ 1 - P( \operatorname{type.order}  \leq  \operatorname{Scab} ) } \right] &= \alpha_{1} + \beta_{1}(\operatorname{class}_{\operatorname{srw}}) + \beta_{2}(\operatorname{density}) + \beta_{3}(\operatorname{hardness}) + \beta_{4}(\operatorname{size}) + \beta_{5}(\operatorname{weight}) + \beta_{6}(\operatorname{moisture}) \\
\log\left[ \frac { P( \operatorname{type.order}  \leq  \operatorname{Sprout} ) }{ 1 - P( \operatorname{type.order}  \leq  \operatorname{Sprout} ) } \right] &= \alpha_{2} + \beta_{1}(\operatorname{class}_{\operatorname{srw}}) + \beta_{2}(\operatorname{density}) + \beta_{3}(\operatorname{hardness}) + \beta_{4}(\operatorname{size}) + \beta_{5}(\operatorname{weight}) + \beta_{6}(\operatorname{moisture})
\end{aligned}
$$
## evaluate

- $\alpha_1 = 17.57$ and $\alpha_2 = 20.04$ and their corresponding t-stat are 7.82 and 8.56 respectively.

```{r}
summary(mod.fit.ord)
```

- density and weight have low `p-value`

```{r}
Anova(mod.fit.ord)
```

## predict $\pi_j$

```{r}
pi.hat.ord <- predict(object = mod.fit.ord , type = "probs")

head(pi.hat.ord)
```

- For finding confidence interval see page 175

## OR

- p175
- The odds of $Y \le j$ vs $Y > j$ change by $e^{c\beta_1}$ times for a `c-unit` increase in $x_1$ while holding the other explanatory variables in the model constant.

- The OR stays the same no matter what response category is used for `j`